---
title: "Airway RNA-seq Differential Expression Analysis"
subtitle: "Dexamethasone-treated vs Untreated Human Airway Smooth Muscle Cells"
author: "Petros Kogios"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 6
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Executive Summary

Bulk RNA-seq differential expression analysis comparing dexamethasone-treated (`trt`)
vs untreated (`untrt`) human airway smooth muscle cells (n=8, 4 cell lines × 2 conditions).
At an adjusted p-value threshold of 0.05, we identified **4,099 differentially expressed
genes** (2,201 upregulated, 1,898 downregulated). The first principal component (48%
variance) cleanly separates treatment groups, and top differentially expressed genes
include known glucocorticoid response genes (PER1, MT2A, SPARCL1), confirming a strong
and biologically coherent transcriptional response.

## Background

Dexamethasone is a synthetic glucocorticoid widely used to treat inflammatory diseases
including asthma. In airway smooth muscle (ASM) cells, glucocorticoids modulate
contractility, proliferation, and inflammatory signaling. Understanding the
transcriptional response of ASM cells to dexamethasone is relevant to mechanisms of
glucocorticoid action in asthma therapy.

### Dataset

We used the `airway` Bioconductor package (Himes et al. 2014, PMID: 24926665):

- **Samples**: 8 RNA-seq libraries
- **Cell lines**: 4 primary human ASM cultures (N052611, N061011, N080611, N61311)
- **Conditions**: untreated (`untrt`) vs 1 µM dexamethasone for 18h (`trt`)
- **Design**: paired (each cell line measured under both conditions)
- **Data type**: gene-level read counts in a `SummarizedExperiment`

```{r setup}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(knitr)

# Load pre-computed results from the analysis pipeline
dds <- readRDS("../data/dds.rds")
res <- readRDS("../data/results.rds")
```

## Methods

### Pre-filtering

To remove genes with very low counts (which provide no statistical power and inflate
the multiple testing burden), we retained only genes with at least 10 reads in 3 or
more samples. This filter reduced the gene set from 63,677 to 16,596 expressed genes.

### Differential expression model

We fit a negative binomial generalized linear model with DESeq2 (v1.46) using the
design formula `~ cell + dex`. Here, `cell` is a batch covariate (4 distinct primary
cultures with different genetic backgrounds), and `dex` is the variable of interest.
By modeling cell line explicitly, the test for `dex` is performed *within-line* and
is robust to inter-individual variability.

We used the Wald test with the contrast `trt vs untrt`, and adjusted p-values for
multiple testing via the Benjamini-Hochberg procedure. Significance threshold:
adjusted p-value < 0.05.

## Results

### Sample-level quality control

```{r pca}
vsd <- vst(dds, blind = FALSE)
pca_data <- plotPCA(vsd, intgroup = c("dex", "cell"), returnData = TRUE)
pv <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, color = dex, shape = cell)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", pv[1], "%")) +
  ylab(paste0("PC2: ", pv[2], "%")) +
  theme_bw() +
  ggtitle("Principal Component Analysis")
```

PC1 (48% of variance) cleanly separates treated from untreated samples, indicating
that the dexamethasone effect is the dominant source of transcriptional variability
in this dataset. PC2 (24%) reflects cell-line differences, which are controlled for
in the design formula.

### Differential expression summary

```{r summary-table}
n_total <- sum(!is.na(res$padj))
n_sig <- sum(res$padj < 0.05, na.rm = TRUE)
n_up <- sum(res$padj < 0.05 & res$log2FoldChange > 0, na.rm = TRUE)
n_down <- sum(res$padj < 0.05 & res$log2FoldChange < 0, na.rm = TRUE)

summary_df <- data.frame(
  Metric = c("Genes tested", "Significant DEGs (padj < 0.05)",
             "Upregulated", "Downregulated"),
  Value = c(n_total, n_sig, n_up, n_down)
)
kable(summary_df)
```

### Volcano plot

```{r volcano}
res_df <- as.data.frame(res)
res_df$significant <- with(res_df,
    !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)

ggplot(res_df, aes(log2FoldChange, -log10(padj), color = significant)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("grey70", "red3")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  theme_bw() +
  labs(title = "Volcano: trt vs untrt",
       x = "log2 fold change", y = "-log10 padj")
```

The volcano shows a symmetric distribution of significant DEGs (red) on both sides
of zero log2 fold change. The most significant hits show -log10(padj) > 100,
consistent with strong, biologically coherent regulation of canonical glucocorticoid
target genes.

### Top 30 differentially expressed genes

```{r heatmap}
#| fig-height: 9
res_df$gene <- rownames(res_df)
res_sig <- res_df[!is.na(res_df$padj) & res_df$padj < 0.05, ]
res_sig <- res_sig[order(res_sig$padj), ]
top_genes <- head(rownames(res_sig), 30)

mat <- assay(vsd)[top_genes, ]
mat <- mat - rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, c("dex", "cell")])

pheatmap(mat, annotation_col = anno, show_rownames = TRUE,
         main = "Top 30 DEGs by padj (VST, row-centered)")
```

The heatmap shows complete separation of treated and untreated samples via
hierarchical clustering, including canonical glucocorticoid-responsive genes such
as **PER1** (ENSG00000179094, circadian clock regulator), **MT2A** (ENSG00000125148,
metallothionein), and **SPARCL1** (ENSG00000152583). The biological coherence of
these top hits provides a strong positive control for the analysis pipeline.

### Top 10 DEGs

```{r top10-table}
top10 <- head(res_sig[, c("log2FoldChange", "padj", "baseMean")], 10)
top10$gene_id <- rownames(top10)
top10 <- top10[, c("gene_id", "log2FoldChange", "padj", "baseMean")]
kable(top10,
      digits = c(0, 2, 50, 1),
      caption = "Top 10 differentially expressed genes (ranked by adjusted p-value)",
      row.names = FALSE)
```

## Limitations

- **Small sample size** (n=4 per condition) limits statistical power for detecting
  small effect sizes.
- **No effect-size shrinkage applied**: log2 fold changes for low-count genes may be
  noisy. A future iteration should apply `lfcShrink()` (apeglm or ashr) for more
  reliable effect-size estimates, particularly for ranking and visualization.
- **No functional enrichment performed**: pathway analysis (GO, KEGG, Reactome) would
  strengthen biological interpretation and identify enriched gene programs.
- **Single time point**: the dataset captures the response at 18h post-treatment;
  early and late dynamics are not resolved.
- **No gene symbol annotation**: results are reported with Ensembl IDs only.
  Mapping to HGNC symbols would improve interpretability.

## Conclusions

The DESeq2 analysis identifies a strong, biologically coherent transcriptional
response to dexamethasone in human airway smooth muscle cells, with 4,099 DEGs at
padj < 0.05. Top-ranked hits include canonical glucocorticoid response genes,
confirming the validity of the analysis pipeline. The clean separation of conditions
on PC1 and via hierarchical clustering of top DEGs indicates a robust and reproducible
treatment effect across all four cell lines.

## Session Info

```{r session-info}
sessionInfo()
```

## References

Himes BE, Jiang X, Wagner P, Hu R, Wang Q, Klanderman B, et al. *RNA-Seq Transcriptome
Profiling Identifies CRISPLD2 as a Glucocorticoid Responsive Gene that Modulates
Cytokine Function in Airway Smooth Muscle Cells.* PLoS One. 2014;9(6):e99625.
doi:10.1371/journal.pone.0099625
