---
title: "RNA-seq Differential Expression Analysis"
subtitle: "Bioinformatics Agent Studio"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 6
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(knitr)
library(jsonlite)

dds <- readRDS("../data/dds.rds")
res <- readRDS("../data/results.rds")
summary_data <- fromJSON("../data/analysis_summary.json")

res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)
res_sig <- res_df[!is.na(res_df$padj) & res_df$padj < 0.05, ]
res_sig <- res_sig[order(res_sig$padj), ]

vsd <- tryCatch(
  vst(dds, blind = FALSE),
  error = function(e) varianceStabilizingTransformation(dds, blind = FALSE)
)

# Auto-detect intgroup columns (factor/character only, excluding sizeFactor)
all_cols <- colnames(colData(dds))
discrete_cols <- all_cols[sapply(colData(dds), function(x) is.factor(x) || is.character(x))]
intgroup_cols <- setdiff(discrete_cols, c("sizeFactor", "replaceable"))
if (length(intgroup_cols) == 0) intgroup_cols <- all_cols[1]

color_var <- intgroup_cols[1]
shape_var <- if (length(intgroup_cols) >= 2) intgroup_cols[2] else NULL

# Detect optional outputs
pathway_img <- "../results/figures/04_pathway_enrichment.png"
pathway_dotplot <- "../results/figures/05_top_pathways_dotplot.png"
has_pathway <- file.exists(pathway_img)
heatmap_img <- "../results/figures/03_heatmap_top30.png"
has_heatmap <- file.exists(heatmap_img)
```

{{< include narrative.md >}}

## Dataset

| Field | Value |
|---|---|
| Comparison | `r summary_data$comparison` |
| Organism | `r summary_data$organism` |
| Total samples | `r ncol(dds)` |
| Genes tested (after filtering) | `r summary_data$genes_after_filter` |
| Design formula | `r summary_data$design_formula` |

## Methods

DESeq2 (v1.46) was used to fit a negative binomial generalized linear model with the design formula shown above. Differential expression was assessed using the Wald test with Benjamini-Hochberg adjustment for multiple testing. Significance threshold: padj < 0.05.

## Results

### Sample-level QC (PCA)

```{r pca}
pca_data <- plotPCA(vsd, intgroup = intgroup_cols, returnData = TRUE)
pv <- round(100 * attr(pca_data, "percentVar"))

p <- if (!is.null(shape_var)) {
  ggplot(pca_data, aes_string("PC1", "PC2", color = color_var, shape = shape_var))
} else {
  ggplot(pca_data, aes_string("PC1", "PC2", color = color_var))
}

p + geom_point(size = 4) +
  xlab(paste0("PC1: ", pv[1], "%")) +
  ylab(paste0("PC2: ", pv[2], "%")) +
  theme_bw() +
  ggtitle("Principal Component Analysis")
```

### Summary

```{r summary-table}
summary_df <- data.frame(
  Metric = c("Genes tested", "Significant DEGs (padj < 0.05)",
             "Upregulated", "Downregulated"),
  Value = c(sum(!is.na(res$padj)),
            nrow(res_sig),
            sum(res_sig$log2FoldChange > 0),
            sum(res_sig$log2FoldChange < 0))
)
kable(summary_df)
```

### Volcano plot

```{r volcano}
res_df$significant <- with(res_df,
    !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)
ggplot(res_df, aes(log2FoldChange, -log10(padj), color = significant)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("grey70", "red3")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  theme_bw() +
  labs(title = paste("Volcano:", summary_data$comparison),
       x = "log2 fold change", y = "-log10 padj")
```

### Top DEGs

```{r top-genes}
top <- head(res_sig, 20)
display_cols <- intersect(c("gene", "gene_symbol", "log2FoldChange", "padj", "baseMean"),
                          colnames(top))
kable(top[, display_cols], digits = 4)
```

```{r heatmap-section, eval=has_heatmap, results='asis'}
cat("### Heatmap of top DEGs\n\n")
cat("![](", heatmap_img, ")\n", sep = "")
```

```{r pathway-section, eval=has_pathway, results='asis'}
cat("## Pathway Enrichment\n\n")
cat("Enriched biological pathways and gene sets across GO, KEGG, and Reactome:\n\n")
cat("### Manhattan plot\n\n")
cat("![](", pathway_img, ")\n\n", sep = "")
cat("### Top enriched terms\n\n")
cat("![](", pathway_dotplot, ")\n", sep = "")
```

## Reproducibility

```bash
docker run --rm \
    -v /local/data:/data \
    -v /local/output:/output \
    --env-file .env \
    bioagent-studio \
    --counts /data/counts.csv \
    --metadata /data/metadata.csv \
    --condition <condition_col> \
    --reference <reference_level> \
    --output /output/
```

---

*Report generated by Bioinformatics Agent Studio. Scientific narrative produced by an AI agent (Claude Sonnet) constrained to use only the analysis JSON output as input.*
