---
title: "Airway RNA-seq Differential Expression Analysis"
subtitle: "Dexamethasone-treated vs Untreated Human Airway Smooth Muscle Cells"
author: "Petros Kogios"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    theme: cosmo
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 6
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(knitr)

dds <- readRDS("../data/dds.rds")
res <- readRDS("../data/results.rds")

res_df <- as.data.frame(res)
res_df$gene <- rownames(res_df)
res_sig <- res_df[!is.na(res_df$padj) & res_df$padj < 0.05, ]
res_sig <- res_sig[order(res_sig$padj), ]
vsd <- vst(dds, blind = FALSE)
```

{{< include narrative.md >}}

## Dataset

`airway` Bioconductor package (Himes et al. 2014, PMID: 24926665).

- **Samples**: 8 RNA-seq libraries
- **Cell lines**: 4 primary human ASM cultures (N052611, N061011, N080611, N61311)
- **Conditions**: untreated (`untrt`) vs 1 ÂµM dexamethasone for 18h (`trt`)
- **Design**: paired (each cell line measured under both conditions)

## Methods

We fit a negative binomial generalized linear model with DESeq2 (v1.46) using the
design formula `~ cell + dex`. The `cell` covariate controls for inter-individual
variability across the 4 primary cultures. We used the Wald test with the contrast
`trt vs untrt` and Benjamini-Hochberg correction. Significance threshold: padj < 0.05.

Pre-filtering retained genes with â‰Ą10 reads in â‰Ą3 samples (63,677 â†’ 16,596 genes).

## Results

### Sample-level QC

```{r pca}
pca_data <- plotPCA(vsd, intgroup = c("dex", "cell"), returnData = TRUE)
pv <- round(100 * attr(pca_data, "percentVar"))

ggplot(pca_data, aes(PC1, PC2, color = dex, shape = cell)) +
  geom_point(size = 4) +
  xlab(paste0("PC1: ", pv[1], "%")) +
  ylab(paste0("PC2: ", pv[2], "%")) +
  theme_bw() +
  ggtitle("Principal Component Analysis")
```

### Summary table

```{r summary-table}
summary_df <- data.frame(
  Metric = c("Genes tested", "Significant DEGs (padj < 0.05)",
             "Upregulated", "Downregulated"),
  Value = c(sum(!is.na(res$padj)),
            nrow(res_sig),
            sum(res_sig$log2FoldChange > 0),
            sum(res_sig$log2FoldChange < 0))
)
kable(summary_df)
```

### Volcano plot

```{r volcano}
res_df$significant <- with(res_df,
    !is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1)

ggplot(res_df, aes(log2FoldChange, -log10(padj), color = significant)) +
  geom_point(alpha = 0.6, size = 1) +
  scale_color_manual(values = c("grey70", "red3")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  theme_bw() +
  labs(title = "Volcano: trt vs untrt",
       x = "log2 fold change", y = "-log10 padj")
```

### Heatmap (top 30 DEGs)

```{r heatmap}
#| fig-height: 9
top_genes <- head(rownames(res_sig), 30)
mat <- assay(vsd)[top_genes, ]
mat <- mat - rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, c("dex", "cell")])
pheatmap(mat, annotation_col = anno, show_rownames = TRUE,
         main = "Top 30 DEGs by padj (VST, row-centered)")
```

### Top 10 DEGs

```{r top10-table}
top10 <- head(res_sig[, c("gene", "log2FoldChange", "padj", "baseMean")], 10)
kable(top10, digits = c(0, 2, 50, 1), row.names = FALSE,
      caption = "Top 10 DEGs ranked by adjusted p-value")
```

## Session Info

```{r session-info}
sessionInfo()
```

## Reference

Himes BE et al. *RNA-Seq Transcriptome Profiling Identifies CRISPLD2 as a
Glucocorticoid Responsive Gene that Modulates Cytokine Function in Airway
Smooth Muscle Cells.* PLoS One. 2014;9(6):e99625.
